Analysis

### 2019-08-16 reports/manuscript.Rmd
new version of paper - finished exp2 - finished discussion - half of the introduction

  Work of FD was supported by student grant no. 604218 nad by RVO 68081740. work JL was supported by GACR no. 16-07983S.

@@

@@

knitr::include_graphics(here("plots","Fig_odd_results.png"), here("plots","Fig_odd_results_ratio.png"))

@@

We selected the stimuli for the experiment from the database FIGRIM [@bylinskii2015], which contains 21 categories, each with more than 300 scenes of size 700$\times$700 px.

>>>>

  Work of FD was supported by student grant no. 604218 and by RVO 68081740. work JL was supported by GACR no. 16-07983S.

@@

# Introduction

World around us is full of information to process and store. When styduing visual perception, it has been observed that visual memory has massive capacity for complex stimuli [@standing1970, @standing1973]. Recent studies showed that we are able to store not only shallow representation of the objects, but they are able to store many details [@brady2008] and similar findings were observed for scenes [@konkle2010b]. 
Although the capacity is large, it is an open question, how is it organised. The high recollection accuracy get lower when using yes/no task instead of 2AFC [@Cunningham2015]. Moreover, recent results show that the memory is much more limited in detail than it seems [@Draschkow2019].

How is the long-term memory orgranised? One possible explanation would be to create a coding scheme, using which the memories will be organised [@brady2011], similarly as in working memory [@brady2009]. This coding scheme is created during our lifetime with the formation of semantical categories [@Kemp2009]. In the case of objects, the possible candidate is a hierarchical structure from simple features through larger meaningful groups to whole objects [@brady2001, @Ullman2007]. In the case of scenes is the situation more complicated as not only the objects in the scene should be represented, but their spatial arrangment as well as the gist of the scene should be stored [@Oliva2005]. The organisation of whole scenes in the memory remains an open question.

One possible approach to this problem could be through formation of categories. The semantic categories in our mind are formed to match the categories of outside world and to be economic with terms of used properties to describe the structure [@rosch1978].

@@

knitr::include_graphics(c(here("plots","Fig_odd_results.png"), here("plots","Fig_odd_results_ratio.png")))

@@

We selected the stimuli for the experiment from the database FIGRIM [@bylinskii2015], which contains 21 categories, each with more than 300 scenes of size 700$\times$700 px. Photographs of scenes in FIGRIM database come from larger SUN database [@xiao2010], which contain many artifacts (e.g. some scenes were not photographs of real scenes, but they were built from lego bricks, contain camera recording markers, etc.). Before the selection of the stimuli, we manually removed are scenes that had some artifacts in them. After the selection, 8665 photographs remained. For those scenes, we again evaluated each of them by pre-trained CNN and created an image space from the activation of fc7 layer.

whole design is visualized in Figure \@ref(fig:drm-scheme).For each scene category, we first computed the category center. then we divided the scenes by the distances from the center into groups closer than median and farther than median. From the closer group, we randomly selected 15 scenes, which were used in training block. These 15 scenes create a polygon in the image space and therefore we could compute the its center representing average target scene. From the group of closer scenes, we select 5 more scenes closest to the center of the training scenes. These scenes will de denoted as *close-distractors*. From the group of farther scenes, we select randomly 5 scenes, which will be denoted as *far-distractors*. Finnaly, from the training scenes, we selected five which will be presented in testing block. 
Each category was sampled three times, while unique scenes were used for each set.

(ref::drm-scheme) Schema of experimental design. The black x denotes category center and dotted circle group of closer scenes. For randomly selected targets from the closer group (green dots), we compute the center (green x) and select close-distractors (red dots). From the farther scenes, we select far-distractors (blue dots).

```{r drm-scheme, fig.cap = "(ref:drm-scheme)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_scheme.png"))

Experiment was programmed in PsychoPy. Participants were seated approximately 60 cm from the LCD display (22") Experiment consisted alternating parts consisting of training blocks (15 training scenes) and testing blocks (15 scenes). From the 15 testing scenes, five were previously presented target scenes, five were close-distractors and remaining five were far-distractors. Before each training scene, fixation cross was shown for 300 ms. Before the first scene in experimental training block, name of the category was shown instead. In the testing part, participants responded using keyboard. First, they responded using arrow, whether the scene was presented or not. Left arrow denoted presented scene, while right arrow denoted novel scene. Then they rated their confidence on scale "not sure at all", "somewhat sure", and "very sure". The selection of the confidence was again done using arrow keys (down arrow confirmed the selection)-Whole procedure is visualized in Figure \@ref(fig:drm-procedure).
Participants were instructed that the speed of answer is not important, just the accuracy. There were no breaks in the experiment, but participants were instructed to take breaks when needed. Experiment lasted 50-70 minutes.

(ref::drm-procedure) Experimental procedure for one training and testing blog.

```{r drm-procedure, fig.cap = "(ref:drm-procedure)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_procedure.png"))

### Data analysis

Analysis was again performed in statistical tool R. The differences between three types of scenes (target, close-distractor, far-distractor), with focus on differences between far- and close- distractors, were tested using generalized linear mixed models. Participant id and category id as random factors while keeping the model maximal (both intercept and slope). Similarly as in Experiment 1, we tested the differences against chance level (50%).

We also analyzed the data using Signal detection theory approach [@green1966,@MacmillanNA2004]. Use of confidence ratings allowed us to construct empirical ROC curves with nonunit slopes for discriminating between targets and close-distractors and targets and far-distractors. We expressed the performance using $d_a$ measure as an alternative to traditional $d'$ used in a case of non-unit ROC slopes, which is recommened to be use in memory studies [@Loiotile2015]. Using this approach, the bias is not computed traditionally, but instead the the perceptual distance between scene types is computed [@MacmillanNA2004, @Loiotile2015].

The recognition accuracy for target scenes was 69% (SD = 12%) as visualized in Figure \@ref(fig:drm-results-main). In the case of distractors, the incorrect answer represent a false alarm. The false alarm rate for close distractors was 24% (SD = 17%), while for far-distractors the false alarm rate was only 11%.

One sample t-tests revealed that recognition was above chance level in all three conditions (_p_s < .001, Cohen's _d_s > 1.52). The difference between two distractor types was sinigficant ($\chi^2(1) = 56.1, _p_ < .001$)

(ref::drm-results-main) Accuracy results for each type of scene. The shape of data distribution is visualized using violin plot. The vertical line shows bootstrapped standard error of the mean, horizontal line denotes chance level. The false alarm rate for distractors could be computed by subtracting the accuracy from 100%.

```{r drm-results-main, fig.cap = "(ref:drm-results-main)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_main.png"))

By using the confidence data taken together with old/new answers, we obtained empirical ROC curves (visualized in Figure \@ref(fig:drm-results-ROC)) for both discrimination conditions. Both curves had slope lesser then 1 (targets vs close distractors: slope = 0.62, SD = 0.14; targets vs far distractors: slope = 0.75, SD = 0.23) and the difference from unit slope was significant (_p_s < .001), so we had to use index $d_a$ for measuring sensitivity. Discriminability between target scenes and close distractors was 1.15 (SD = 0.56) and between target scenes and far distractors was 1.71 (SD = 0.61). The difference in sensitivity was significant (paired t-test: t(25) = -14.33, _p_ < .001, _d_ = 2.81). When visualizing the schema SDT model (Figure \@ref(fig:drm-results-SDT-model)), the perceptual distance between distractors [@Dosher1984] was small (pseudo $d_a$ = 0.56).

(ref::drm-results-ROC) ROC curves for discrimination between Target and close-distractors (left) and target and far-distractors (right). Black line denotes chance level. Red line shows average ROC curve, while the grey lines shows ROC curves for individual participants.

```{r drm-results-ROC, fig.cap = "(ref:drm-results-ROC)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_ROC.png"))

(ref::drm-results-SDT-model) SDT model for three scene types. The perceptual distance between distractors was small (pseudo $d_a$ = 0.56).

```{r drm-results-SDT-model, fig.cap = "(ref:drm-results-SDT-model)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_SDT_model.png"))

The exploration of recognition accuracy per each scene category revealed that easiest were categories bridge and bathroom, while most difficult were badlands and cockpit. Full results are displayed in Figure \@ref(fig:drm-results-category). On the level individual scene, we explored, whether the memory accuracy could be explained by memorability of the photograph [@Isola2001]. The memorability is good candidate to predict the task, because it has been shown that memorable images shows good perceptual organisation [@Goetschalckx2018] and are quickly categorisable [@Broers2017]. Therefore, we evaluated the photographs using trained CNN LaMem, which can predict memorability similarly to the behavioral scores [@khosla2015]. We then computed correlation of memory accuracy with this predicted memorability. The results showed that the correlation was small and significant for targets (_r_ = .10, _p_ = .003)) and for close-distractors(_r_ = .15, _p_ = .007)), while not significant for far-distractors (_r_ = .05, _p_ = .335). Therefore, the effect of memorability is only small.

(ref::drm-results-category) Moving average for reaction time (size of window was 200 consescutive trials) with respect to id of the trial.

```{r drm-results-category, fig.cap = "(ref:drm-results-category)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_categories.png"))

We were also interested, whether the recognition of the presented scenes propagated to the time needed to answer. We excluded the responses longer than 5s. Participants tended to make faster responses during the time course of the experiment as visualized in Figure \@ref(fig:drm-results-rt_window). As visualized in Figure \@ref(fig:drm-results-rt), the fastest responses were for far-distractors (mean = 1.09 s, SD = 0.29), followed by target scenes (mean = 1.19 s, SD = 0.21), while slowest responses were for close-distractors (mean = 1.26 s, SD = 0.21). The differences were significant ($\chi^2(1) = 27.8, _p_ < .001$). Pairwise post-hoc analysis with Tukey's correction revealed that the difference in reaction times between targets and close-distractors were not significant (_p_ = .256), while the remaing pairwise differences were signifcant (p < .001). Moreover, the difference in reaction times between close- and far-distractors was large (Cohen's d = 1.57).

(ref::drm-results-rt-window) Moving average for reaction time (size of window was 200 consescutive trials) with respect to id of the trial.

```{r drm-results-rt-window, fig.cap = "(ref:drm-results-rt-window)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_rt_window.png"))

(ref::drm-results-rt) Reaction times for each scene type. The shape of data distribution is visualized using violin plot. The vertical line shows bootstrapped standard error of the mean.

```{r drm-results-rt, fig.cap = "(ref:drm-results-rt)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_rt.png"))

Because the experiment was long and potentially tireing, we analyzed the accuracy using moving average similarly as for reaction times. As visualized in Figure \@ref(fig:drm-results-window-acc), the accuracy remained stable for distreactors and declined for targets after approximately 1000th trial. However, the critical difference between distractors was stable across the whole trial.

(ref::drm-results-window-acc) Moving average (N = 200) for recognition accuracy. Accuracy for the distractors was stable, while the accuracy declined approximately after 1000th trial.

```{r drm-results-window-acc, fig.cap = "(ref:drm-results-window-acc)", fig.align = 'center'}
knitr::include_graphics(here("plots","Fig_drm_results_rt.png"))

In this experiment, we explored, whether we form a perceptual prototype when memorizing complex stimuli such as scenes. We found out that distractors closer to the average of the presented scenes show higher false alarm rate than distractors far from the average. Similar results were observed for simpler stimuli such as geometric stimuli [@solso1979] or simple faces [@Solso1981], not for complex scenes.

Here presented evidence about the prototypical formation of scene average is not definite proof about use of image space as a model of memory for scenes. Recent literature however shows that in a terms of scene classification, deep features show high resemblance to functional connectivity [@Groen2018]. Moreever the presented approach could be easily extended to compare between several competing models. We could select stimuli in a similar manner using different image spaces and see, which structure would show more false alarms.

Analysis of reaction times showed that scenes closer to the visual prototype required more time to respond. This is in line with findings regarding reaction time in prototype formation for numerical prototypes [@Solso1993] or for scene classification using adjectives [@greene2009].

One problem with image space that as the representation is highly dimensional, it is very densly populated even for large scene databases. It limits finer experimental manipulation in order to create actual visual prototype. One possible solution to this problem would be to use either 3d models of the scene or take advantage of recent advances in artificial scenes created by Generative adversial networks [@goodfellow2014] which were recently prototyped for scene creation based on the semantic description [@park2019].

Our reserach complements findings from scene categorization, which show similar resemblance to scene memory. Research focused on scene categorization showed converging evidence that categorizing scenes by their function is best model explaining human performance and it outperforms deep features [@greene2015]. However on the neural level, the functional connectivity appears as better model showing higher resemblance with behavioral data [@Groen2018]. Similar findings could be observed in a case of visual memory for the scenes.

# Conclusion

In these two experiments, we explored the possible use of convolutional neural networks as a possible model for representation of the memory. We observed that participants intuitively understand the distances in the image space (Experiment 1) and when we preselect stimuli based on the their distance as estimated by deep features, we can observe typical effects in formation of prototypes. Taken together with our past research using deep features [@lukavsky2017], we believe that measure of perceptual similarity as defined by convolutional neural networks shows promising result as a valid metric.

# Acknowledgement

The work of F.D. on this research was supported by a student grant GAUK No. 604218 and by Grant No. RVO 68081740. The work of J.L. was part of a research program of the Czech Academy of Sciences, Strategy AV21, and was supported by Czech Science Grant Foundation No. 16-07983S.
